Example Pipeline for BigSmallDNA

<!------------------------------------------------------------------------------------------------>

# Example pipeline details

Hardware Details:

- Model: Dell Inc. Precision 7730
- CPU: Intel Core i7-8750H × 12
- RAM: 32 GB
- VRAM: 8 GB
- OS: Fedora Linux 40 (Workstation) with Linux 6.11.4-201.fc40.x86_64

Softwares Details:

- R 4.4.1
    - tidyverse 2.0.0
    - stringr 1.5.1
    - kmer 1.1.2
    - ape 5.8.1
    - pegas 1.3
    - dendextend 1.19.0
- GNU bash 5.2.26
- datasets 16.22.1
- SeqKit 2.8.2
- EMBOSS 6.6.0.0
- MAFFT 7.526

Other Resources:

- NCBI Web Server

<!------------------------------------------------------------------------------------------------>

# Instructions

1. Filter genomes on NCBI using the following criteria and download the csv results table:
    - Taxon ID: 2697049 (CoV-2)
    - Ambiguous Characters < 30
    - Nucleotide Completeness: Complete
    - Host: 9696 (human)
    - Collection date: 30th December 2003 - current date
2. Extract accession IDs from the results tables.
3. Download genomes using NCBI datasets and remove duplicates using SeqKit.
4. Extract ORFs for S gene using EMBOSS and remove duplicates using SeqKit.
5. Cluster the sequences using h-clustering.
6. Align clusters roughly using MAFFT and remove duplicates using SeqKit.
7. Improve genetic diversity for each lineage.

<!------------------------------------------------------------------------------------------------>

# Pipeline

```{r setup, include=FALSE}

# disable rendering of chunks
knitr::opts_chunk$set(eval = FALSE)
knitr::opts_chunk$set(include = FALSE)

# working directory
wd <- "data"

# libraries
library(tidyverse) # general data manipulation and visualization
library(stringr) # string manipulation
library(ape) # read, store (DNAbin), and write fasta
library(pegas) # nuc.div()
library(kmer) # kdistance()
library(dendextend) # sub-clustering

```

```{bash wd setup}

mkdir data

# data structure
cd $wd
mkdir $wd/raw fasta msa

```

```{bash 1: filter and download results table}

# open NCBI Virus
xdg-open "https://www.ncbi.nlm.nih.gov/labs/virus/vssi/#/"

# search 2697049
# enter the mentioned inclusion criteria
# download csv result table as "data/raw/res.csv" with the following: Accession, Pangolin, with version.

```

```{r 2: extract ids}

## read csv results table
res <- read.csv(paste0(wd, "/raw/res.csv"))

# make index for writing IDS
temp <- c(seq(1, nrow(res), 10000), nrow(res))

# write IDs
for (i in 1:length(temp)) {
    write.table(
        res$Accession[temp[i-1]:temp[i]],
        paste0(wd, "/raw/ID.", i-1, ".txt"),
        row.names = FALSE, col.names = FALSE, quote = FALSE
        )
    }


```

```{bash 3: download sequences}

# wd
cd $wdDatasets

# assuming 100 files of IDs, so change index accordingly

# download sequences
for i in {1..136}
do
    ./datasets download genome accession --inputfile $wd/raw/ID.$i.txt --filename $wd/raw/ID.$i.zip
done

# extract
for i in {1..136}
do
    unzip $wd/raw/ID.$i.zip -d $wd/raw/ID.$i
done

# empty file to merge all seqs
> $wd/raw/dataset

# add all seqs to empty file
for i in {1..136}
do 
    cat "$wd/raw/res.$i/ncbi_dataset/data/genomic.fna" >> "$wd/raw/dataset"
done

# remove duplicates
seqkit rmdup -s < $wd/raw/dataset > $wd/raw/dataset.fasta

# remove unwanted files
for i in {1..136}
do 
    rm $wd/raw/res.$i.zip # downloaded zip files
    rm $wd/raw/res.$i # extracted folders
    rm $wd/raw/dataset # merged seqs with duplicates
done

```

```{bash 4: extract ORFs}

# wd
cd $wd/raw

# avg length of S gene is 3822 and we keep +-5% margin
getorf -sequence dataset.fasta -outseq ORF -minsize 3631 -maxsize 3860 -find 3

# remove duplicates
seqkit rmdup -s < ORF > ORF.fasta

```

```{r 5: cluster}

# import ORFs
ORF <- read.FASTA(paste0(wd, "/raw/ORF.fasta"))

# CoV-2 has Pangolin lineage so it will be clustered by the classification but the lineages with more than 5000 sequences will be clustered.

# trim labels for ease
labels(ORF) <- substr(labels(ORF), 1, 10)

# trim results table
res <- res[which(res$Accession %in% labels(ORF)),]

# seq frequency of lineages
lin <- data.frame(table((res$Pangolin)))

# group lineages with less than 3 seqs since they can not be clustered
cluster.x <- ORF[labels(ORF) %in% res$Accession[res$Pangolin %in% lin$Var1[lin$Freq <= 3]]]

# determine which lineages do not need to be sub-clustered
temp <- as.character(lin$Var1[lin$Freq > 3 & lin$Freq <= 5000])

# write clusters which do not need to be sub-clustered
for (i in 1:length(temp)) {
    
    print(i) # progress
    
    # write
    write.FASTA(
        ORF[labels(ORF) %in% res$Accession[res$Pangolin %in% temp[i]]],
        file = paste0(wd, "/fasta/cluster.", temp[i], ".fasta")
    )
}

# number of clusters (will be used as index)
var <- 1:length(temp)

# determine which lineages need to be sub-clustered
temp <- as.character(lin$Var1[lin$Freq > 5000])

# repeat the following for all clusters
# not placed in loop because number of clusters need to be checked

# index for cluster
no = (1:length(temp))[1]

# store cluster
seq <- ORF[labels(ORF) %in% res$Accession[res$Pangolin %in% temp[no]]]

# kmean clustering
kmean <- cluster(seq)

# sub-cluster (adjust k accordingly)
cluster <- cutree(as.dendrogram(kmean), k = 1)
View(table(cluster))

# convert to table
cluster <- data.frame(id = names(cluster), cluster = cluster)

# write sub-clusters
for (i in 1:length(unique(cluster$cluster))) {
    
    print(i) # progress
    
    # write
    write.FASTA(
        seq[filter(cluster, cluster == i)$id],
        paste0(wd, "/fasta/cluster.", temp[no], ".", i, ".fasta"),
        )
    
    # update index
    c.var <- c(c.var, paste0(temp[no], ".", i))
    
}

# number of clusters used
# AY.103 - 7
# B.1.1.7 - 17
# B.1.2 - 4
# BA.1.1 - 15
# BA.2.12.1 - 5
# BA.2 - 3

# remove objects and clear cache
rm(list = c("kmean", "cluster", "seq", "no"))
gc()

```

```{bash 6: MSA}

# wd
cd $wd

# MSA
for i in "fasta"/*".fasta"
do
     mafft --retree 1 --thread 12 $i > msa/$(basename $i .fasta).msa.fasta
done

```

```{r 7: genetic diversity}

# read MSA
for (i in 1:length(var)) {
    assign(
        paste0("cluster.", var[i]),
        read.FASTA(paste0(wd, "/msa/cluster.", var[i], ".msa.fasta"))
        )
    }

# dataframe to summarize genetic diversity improvements
nd <- data.frame(
    cluster = paste0("cluster.", var), # cluster
    old.nd = NA, new.nd = NA, change.nd = NA, # old, new, and change in nuc div
    old.seq = NA, new.seq = NA, change.seq = NA, # old, new, and change in num of seq
    th = NA, # threshold
    time = NA # time taken
    )


# improve diversity
for (i in 1:length(.var)) { # check index before running
    
    print(i) # progress 
    
    print(Sys.time()) # time of start
    
    try({ # ignore errors
    
    time.old <- Sys.time() # time for time complexity
    
    # get cluster
    temp <- get(paste0("cluster.", c.var[i])) # run for i>=3
    
    # convert to distance matrix
    ## choose model according to data
    ## current: K80 (default)
    dist <- dist.dna(temp, model = "K80", as.matrix = TRUE)
    # convert diagonal entries to NA for ease
    diag(dist) <- NA
    
    # temporary dataframe to evaluate optimal threshold
    ## chose threshold range according to data
    ## current: 0.9 - 1
    ## if unsure use c(0.5, 0.6, 0.7, 0.8, 0.9, 1) to get an idea
    threshold <- data.frame(th = c(0.9, 0.95, 0.99, 0.999, 0.9999, 1), nd = NA)
    
    # nd of temp (calc once to prevent recalculations)
    temp.nd <- nuc.div(temp)
    
    # evaluate optimal threshold
    # new = nd of seqs with identity >= threshold * max score
    # nd = new - old
    for (j in 1:nrow(threshold)) {
        threshold$nd[j] <-
            nuc.div(
                temp[
                    unique(c(which(
                        dist >= threshold$th[j]*max(dist, na.rm = TRUE),
                        arr.ind = TRUE)))
                    ]
                ) 
        - temp.nd
    }
    
    # determine optimal threshold and save seqs
    assign(
        paste0("cluster", var[i], ".i"),
        temp[
            unique(c(which(
                dist >= threshold$th[
                    order(threshold$nd, decreasing = TRUE)[1]
                    ]*max(dist, na.rm = TRUE),
                arr.ind = TRUE)))
            ]
        )
    
    # nd
    nd$old.nd[i] <- temp.nd
    nd$new.nd[i] <- max(threshold$nd) + temp.nd
    nd$change.nd[i] <- (nd$new.nd[i] - nd$old.nd[i]) / nd$old.nd[i]
    nd$th[i] <- threshold$th[order(threshold$nd, decreasing = TRUE)[1]]
    nd$time[i] <- as.numeric(Sys.time() - time.old)
    nd$old.seq[i] <- length(temp)
    nd$new.seq[i] <- length(get(paste0(var[i], ".i")))
    nd$change.seq[i] <- (nd$new.seq[i] - nd$old.seq[i]) / nd$old.seq[i]
    
    })
    
}

# check if nd was decreased for any cluster
View(nd[which(nd$change.nd <= 0),])

# For negative cases, lower the threshold range and repeat the loop above

# merge clusters
temp <- c()
for (i in 1:length(var)) {
    temp <- c(temp, get(paste0("cluster", var[i], ".i")))
}

# trim results table
res <- res[res$Accession %in% labels(temp),]

# get un-aligned seqs
dataset <- c(ORF[labels(ORF) %in% res$Accession])

# write
write.FASTA(c12, file = paste0(wd, "/fasta/dataset.fasta"))

# remove objects and clear cache
rm(list=c("temp", paste0("cluster.", var), paste0("cluster", var[i], ".i"), "var"))
gc()

```

<!------------------------------------------------------------------------------------------------>
